---
title: "Fluoxetine microbiome analyses"
author: "Gabriel Melhado"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SETUP

## Libraries
### Load all packages required for data import, manipulation, microbiome processing, statistical modelling, visualisation, and file management.

```{r}
# Load packages
pacman::p_load(readr, readxl, here, dplyr, tidyr, tidyverse, forcats, scales, RColorBrewer, grid, janitor, stringr, tibble, phyloseq, vegan, microbiome, microViz, ggside, pairwiseAdonis, DESeq2, ggplot2, lme4, lmerTest, glmmTMB, patchwork, Maaslin2, BiocManager)

```


# Data

## Data wrangling
### Import the experimental metadata and format sample identifiers so they match the identifiers used in the sequencing files. The sample ID column is then converted to row names and stored as a sample_data object for integration with phyloseq.

```{r}
# Read and prepare metadata
meta <- read_csv(here("data", "experimental_data.csv"))


# Add .fastq to sample column so that it will match the sequence data and make them row names
meta$sample_id <- paste(meta$sample_id, ".fastq", sep = "")
meta <- meta %>%
  remove_rownames() %>%
  column_to_rownames(var = "sample_id")

# Create phyloseq object                
mtd_final <- sample_data(meta)

```

## Sequencing information
### Import the ASV abundance table, taxonomic assignments, and rooted phylogenetic tree produced by the sequencing pipeline. Two alternative taxonomic-classification outputs are imported here so their results can be compared before selecting one for downstream analyses.

```{r}
# Feature table (ASVs count and taxonomy info using Naive Bayes classifier)
biom_NB <- import_biom(
  here("data", "feature-table-tax.biom"),
  treefilename = here("data", "phylotree_mafft_rooted.nwk")
)

# Feature table (ASVs count and taxonomy info using VSEARCH)
biom_VSEARCH <- import_biom(
  here("data", "feature-table-tax_vsearch.biom"),
  treefilename = here("data", "phylotree_mafft_rooted.nwk")
)

```

# Generate phyloseq objects

### Note that the Naive Bayes and VSEARCH classifications returned similar ASV-level outputs/taxonomic patterns. We therefore proceeded with the Naive Bayes classification for downstream analyses.

## ps_tss will be used for plots and alpha/beta diversity.
## ps_NB includes raw counts and will be used for differential abundance analysis.

# Phyloseq object

### Combine the sequencing data with the experimental metadata to create the initial phyloseq object. Taxonomic rank names are standardised, and a total-sum-scaled copy is produced by converting each sample's counts to relative abundances.

```{r}
# Merge feature table with metadata to create phyloseq object
ps_NB_raw <- merge_phyloseq(mtd_final, biom_NB)
# Fix tax_table so that the column names are correctly names
colnames(tax_table(ps_NB_raw)) <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")

# Total sum scaling: converts raw counts into proportions
ps_tss_raw <- transform_sample_counts(ps_NB_raw, function(x){if(sum(x) == 0){x} # avoid producing naN when the sample count is zero
  else {x / sum(x)}})

```

# Data filtering

## Taxonomy-based filtering
### Apply taxonomy-based quality control to both the relative-abundance and raw-count phyloseq objects. Features lacking phylum-level classification or assigned to chloroplasts, mitochondria, or Chlorophyta are removed because they are not part of the target bacterial community.


```{r}
# Remove features without a phylum-level annotation and those assigned as chloroplast or mitochondria.

# Filter datasets
ps_tss <- ps_tss_raw %>%
  subset_taxa(!is.na(Phylum) & !Phylum %in% c("", "uncharacterised")) %>%
  subset_taxa(Order != "o__Chloroplast" | is.na(Order)) %>%
  subset_taxa(Family != "f__Mitochondria" | is.na(Family)) %>%
  subset_taxa(Phylum != "p__Chlorophyta")

ps_NB <- ps_NB_raw %>%
  subset_taxa(!is.na(Phylum) & !Phylum %in% c("", "uncharacterised")) %>% 
  subset_taxa(Order != "o__Chloroplast" | is.na(Order)) %>% 
  subset_taxa(Family != "f__Mitochondria" | is.na(Family)) %>% 
  subset_taxa(Phylum != "p__Chlorophyta")

```

## Prevalence-based filtering
### Features that show up in only a few samples may not represent real biological diversity but rather PCR/sequencing errors (such as PCR chimeras) or reagent contaminants.
### Remove low-prevalence ASVs that occur in fewer than 10% of samples. This step reduces the influence of extremely rare features that are more likely to reflect sequencing artefacts, contaminants, or poorly supported biological signals.

```{r}
# Prevalence filter (remove features present in less than 10% of samples)
prev_cut <- ceiling(0.10 * nsamples(ps_NB))
ps_NB <- filter_taxa(ps_NB, function(x) sum(x > 0) >= prev_cut, prune = TRUE)

```


### Check filtering
### Compare the number of ASVs before and after filtering. This provides a direct check of how strongly the taxonomic and prevalence filters reduced the feature table.

```{r}
# Check the number of taxa that remain after filtering
ntaxa(ps_NB_raw) 
ntaxa(ps_NB) 

```

```{r}
# Check that no samples have been removed
nsamples(ps_NB_raw)
nsamples(ps_NB) 
```

```{r}
# Check the number of reads that were left after filtering
sum(taxa_sums(ps_NB)) / sum(taxa_sums(ps_NB_raw)) * 100

```

```{r}
summary(sample_sums(ps_NB))
mean(sample_sums(ps_NB))
range(sample_sums(ps_NB))
sd(sample_sums(ps_NB))

```
### Identify samples with fewer than 1,000 retained reads, remove them from the dataset, and then prune ASVs that become absent after sample removal. The sequencing-depth distribution is recalculated to confirm the properties of the final filtered object.


```{r}
# Check sequencing depth per sample
sample_depth <- sample_sums(ps_NB)

# Identify low-depth samples
low_depth_samples <- names(sample_depth[sample_depth < 1000])
low_depth_samples

# Remove samples with <1000 reads
ps_NB <- prune_samples(sample_depth >= 1000, ps_NB)

# Remove taxa/ASVs that are now absent after sample filtering
ps_NB <- prune_taxa(taxa_sums(ps_NB) > 0, ps_NB)

# Check new sequencing depth summary
summary(sample_sums(ps_NB))
mean(sample_sums(ps_NB))
range(sample_sums(ps_NB))
sd(sample_sums(ps_NB))
```
### Visualise ASV prevalence before and after filtering. The histograms show how many samples each ASV occurs in and help confirm that the low-prevalence portion of the distribution has been removed.

```{r}
# Plot histograms to ensure that early peaks have been removed
apply(otu_table(ps_NB_raw), 1, function(x) sum(x > 0)) %>% 
  hist(breaks = 50)
apply(otu_table(ps_NB), 1, function(x) sum(x > 0)) %>% 
  hist(breaks = 50)
```


# Alpha diversity

"Alpha-diversity is the diversity in a single ecosystem or sample. The simplest measure is richness, the number of species (or OTUs) observed in the sample. Other metrics consider the abundances (frequencies) of the OTUs, for example to give lower weight to lower-abundance OTUs." - Robert Edgar

Essentially, alpha diversity measures within-sample diversity across treatments. It is used to evaluate richness and abundance of microbial communities. 

Initial plot: Use the phyloseq library to investigate the alpha diversity measure of each sample across the three treatment groups. Note that we will only use a subset of these measures (explained in further detail below).

## Initial plot

```{r}


plot_richness(ps_NB, x="treatment", color="treatment",
              measures = c("Observed","Shannon","Simpson"))


```

## Calculate alpha diversity

There are many different measures of alpha diversity:
- Observed: The actual number of unique ASVs in a sample. Measures taxonomic richness.
- ACE: Abundance-based coverage estimator. Based on the abundance of species and estimates the no. of additional species that are likely to be present but have not been observed.
- Shannon: Takes into account both species richness and species evenness. It is calculated based on the logarithm of the relative abundances of species in a community.
- Simpson's Diversity Index: Takes into account the number of species present, as well as the relative abundance of each species. 1 = infinite diversity, 0 = no diversity. Measures dominance (i.e., evenness)
- Fisher: Takes into account both the number of species (species richness) and the evenness of their distribution.
- Chao1: Estimates the true number of species (species richness) in a community or sample, especially when some species are rare and may not be observed in the sample. It takes into account both the number of observed species and the number of rare or singleton species to provide an estimate of total species richness.

### Calculate sample-level alpha-diversity metrics and merge them with the experimental metadata. Sample identifiers are standardised before merging so that each diversity estimate is correctly linked to its treatment, tank, and other covariates.

```{r}
# Get alpha values for each sample
alpha <- estimate_richness(ps_NB) %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample") %>%
  mutate(sample = as.character(gsub("X", "", sample)))
# Produces a warning about the lack of singletons (ASVs with a count of 1). This is OK given that the data has been denoised by AGRF, which removes singletons. Note that Chao1 should not be used as it uses singletons in its formula.

# Make sure sample IDs are included in the metadata
IDs <- meta %>%
  rownames_to_column("sample")

# Combine metadata with alpha values
alpha_mtd <- merge(x = IDs, y = alpha, by = "sample")

```

## Analysis:
Is there evidence that alpha values differ across treatment groups? Note that we only report results from Observed, Shannon, and Simpson measures of diversity.

###Test whether alpha diversity differs among fluoxetine treatments while accounting for sequencing depth and tank-level non-independence. Observed richness is modelled with a negative-binomial GLMM, whereas Shannon and Simpson diversity are analysed with linear mixed-effects models.

```{r}
#Tank-level replication
# add sequencing depth
alpha_mtd$reads <- sample_sums(ps_NB)[alpha_mtd$sample]
alpha_mtd$log_reads <- log(alpha_mtd$reads)


#Observed
m_observed <- glmmTMB(
  Observed ~ treatment + log_reads + (1 | tank),
  data = alpha_mtd,
  family = nbinom2()
)

summary(m_observed)

#Shannon
m_shannon <- lmer(
  Shannon ~ treatment + log_reads + (1 | tank),
  data = alpha_mtd
)

anova(m_shannon)
summary(m_shannon)

#Simpson
m_simpson <- lmer(
  Simpson ~ treatment + log_reads + (1 | tank),
  data = alpha_mtd
)

anova(m_simpson)
summary(m_simpson)

```


### Prepare alpha-diversity data for visualisation

```{r}

# Reorder treatment levels
alpha_mtd <- alpha_mtd %>%
  mutate(
    treatment = factor(
      treatment,
      levels = c("C", "L", "H"),
      labels = c("Control", "Low", "High")
    )
  )

# Colour palettes
pastel_cols <- c(
  "Control" = "#A8D5BA",
  "Low"     = "#F9E79F",
  "High"    = "#F5B7B1"
)

dark_cols <- c(
  "Control" = "#2E8B57",
  "Low"     = "#C9A227",
  "High"    = "#B22222"
)

# Long format for plotting raw values
alpha_long <- alpha_mtd %>%
  tidyr::pivot_longer(
    cols = c("Observed", "Shannon", "Simpson"),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, levels = c("Observed", "Shannon", "Simpson"))
  )

# Summary means and 95% CIs for visualisation
alpha_summary <- alpha_long %>%
  group_by(metric, treatment) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / sqrt(sum(!is.na(value))),
    lower = mean - 1.96 * se,
    upper = mean + 1.96 * se,
    .groups = "drop"
  )

# Plot
p_alpha <- ggplot() +
  geom_violin(
    data = alpha_long,
    aes(x = treatment, y = value, fill = treatment),
    alpha = 0.20,
    colour = NA,
    trim = TRUE
  ) +
  geom_jitter(
    data = alpha_long,
    aes(x = treatment, y = value, colour = treatment),
    width = 0.15,
    alpha = 0.60,
    size = 2
  ) +
  geom_errorbar(
    data = alpha_summary,
    aes(
      x = treatment,
      ymin = lower,
      ymax = upper,
      colour = treatment
    ),
    width = 0.25,
    linewidth = 1.2
  ) +
  geom_point(
    data = alpha_summary,
    aes(
      x = treatment,
      y = mean,
      fill = treatment
    ),
    shape = 21,
    size = 5,
    stroke = 1.2,
    colour = "black"
  ) +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = pastel_cols) +
  scale_colour_manual(values = dark_cols) +
  theme_bw(base_size = 14) +
  labs(
    x = "Fluoxetine treatment",
    y = ""
  ) +
  theme(
    strip.background = element_rect(fill = "grey90", linewidth = 1),
    strip.text = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 15),
    panel.grid = element_blank(),
    legend.position = "none",
    panel.border = element_rect(colour = "black", linewidth = 1),
    panel.spacing = unit(1.2, "lines")
  )

print(p_alpha)

ggsave(
  filename = here::here("figures", "alpha_diversity.pdf"),
  plot = p_alpha,
  width = 9,
  height = 4.5
)

ggsave(
  filename = here::here("figures", "alpha_diversity.jpg"),
  plot = p_alpha,
  width = 9,
  height = 4.5,
  dpi = 600
)
```

# Beta diversity

Measures community differences between treatments.

## Setup

First, we'll need to transform the counts so that we account for the number of sequences. Values will no longer be counts, but rather the dominance for each taxa relative to the geometric mean of all taxa on the logarithmic scale

### Apply a centred log-ratio transformation to the filtered ASV count data. This transformation treats the microbiome data as compositional and produces values suitable for Euclidean-distance-based ordination and beta-diversity analysis.

```{r}
# CLR transformation
ps_clr <- microbiome::transform(ps_NB, "clr")
# The values are now no longer counts, but a value of dominance for each taxa

```

Next, conduct the Principle Component Analysis (PCA), examine the relative importance of each PC, and generate the ordination.

### Perform a principal-component ordination of the CLR-transformed community data and display a scree plot. The scree plot shows the proportion of total community variation represented by each ordination axis.

```{r}
# Generate the principal components
ord_clr <- ordinate(ps_clr, "RDA")

# Plot to examine the proportion of total variation explained by each PC
plot_scree(ord_clr) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "\nAxis", y = "Proportion of variance\n")
```

```{r}
# Examine the Eigen values () and the the proportion of variance explained
head(ord_clr$CA$eig)
sapply(ord_clr$CA$eig[1:5], function(x) x / sum(ord_clr$CA$eig))   
```

## Plot ordination
### Calculate the variance explained by the first two ordination axes. These proportions are subsequently used to preserve the appropriate aspect ratio in the PCA plots.

```{r}
# Plot the first two components
# We'll scale the plot to reflect the relative amount of information explained by each axis
clr1 <- ord_clr$CA$eig[1] / sum(ord_clr$CA$eig)
clr2 <- ord_clr$CA$eig[2] / sum(ord_clr$CA$eig)
```

### Create an exploratory PCA plot with treatment-specific points, 95% confidence ellipses, and marginal density distributions. 

```{r}
png("./PCA_temp_1.png", width = 300, height = 225, units='mm', res = 300)
plot_ordination(ps_NB, ord_clr, type = "samples") +
  coord_fixed(clr1/clr2) +
  stat_ellipse(aes(group = treatment, colour = treatment, fill = treatment), 
               geom = "polygon",
               alpha = 0.1,
               linetype = 1, 
               linewidth = 0.8,
               level = 0.95) +
    geom_point(aes(fill = treatment), 
             size = 2, 
             alpha = 0.8, 
             shape = 21) +
  scale_fill_manual(values = c("C" = "#7FA6B8", 
                               "L" = "#F3AD7A", 
                               "H" = "#EC6C31")) +
  scale_colour_manual(values = c("C" = "#7FA6B8", 
                                 "L" = "#F3AD7A", 
                                 "H" = "#EC6C31")) +
  theme_bw(base_size = 14) +
  geom_xsidedensity(aes(fill = treatment),
                    alpha = 0.5,
                    show.legend = FALSE) +
  geom_ysidedensity(aes(fill = treatment),
                    alpha = 0.5,
                    show.legend = FALSE) +
#  geom_xsideboxplot(aes(fill = treatment, y = treatment),
#                    orientation = "y") +
#  geom_ysideboxplot(aes(fill = treatment, x = treatment),
#                    orientation = "x") +
#  scale_xsidey_discrete(labels = NULL) +
#  scale_ysidex_discrete(labels = NULL) +
  theme_ggside_void()
dev.off()
```

```{r}
ord_df <- plot_ordination(ps_NB, ord_clr, type = "samples", justDF = TRUE)
png("./PCA_temp_2.png", width = 300, height = 225, units='mm', res = 300)
ggplot(ord_df, aes(x = PC1, y = PC2)) +
  coord_fixed(clr1/clr2) +
    stat_ellipse(aes(group = treatment, colour = treatment), 
               alpha = 1,
               linetype = 1, 
               linewidth = 1,
               level = 0.95) +
      geom_point(aes(colour = treatment), 
             size = 5, 
             alpha = 0.8) +
  scale_fill_manual(values = c("C" = "#7FA6B8", 
                               "L" = "#F3AD7A", 
                               "H" = "#EC6C31")) +
  scale_colour_manual(values = c("C" = "#7FA6B8", 
                                 "L" = "#F3AD7A", 
                                 "H" = "#EC6C31")) +
  theme_bw(base_size = 14) +
  theme(legend.position = "right",
        legend.title = element_blank(),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 18),
        axis.line = element_line(color = "black", 
                                 linewidth = 0.6),
        panel.border = element_rect(fill = NA, 
                                    color = "black", 
                                    linewidth = 1.5),
        axis.ticks.length = unit(0.25, "cm"),
        legend.text = element_text(size = 18)) +
  xlab(paste0("PC1 (", round(clr1*100, 1), "%)")) +
  ylab(paste0("PC2 (", round(clr2*100, 1), "%)"))
dev.off()
```

It looks like there might be some separate clustering, suggesting possible differences in microbial communities according to treatment. We'll use a permutational multivariate analysis of variance (PERMANOVA) to test whether the samples cluster beyond that expected. Note that PERMANOVA is analogous to ANOVA, but is for multivariate distance data.

## Permanova
### Calculate Euclidean distances among samples using the CLR-transformed ASV table and test for treatment-related differences in community composition with PERMANOVA. This is equivalent to testing Aitchison distances because Euclidean distance is applied after CLR transformation.

```{r}
# Generate Euclidean distance matrix on the clr-transformed ASV data
clr_dist_matrix <- phyloseq::distance(ps_clr, method = "euclidean") 

# Create metadata frame as pairwise.adonis2 doesn't accept vectors
meta <- data.frame(sample_data(ps_clr)) 

# PERMANOVA (ADONIS) test
adonis2(clr_dist_matrix ~ treatment,
        data = meta)
#adonis2(clr_dist_matrix ~ sample_data(ps_clr)$treatment)
# formula = clr_dist_matrix ~ sample_data(ps_clr)$treatment
```

### Prepare the relative-abundance phyloseq object for taxonomic-composition plots and define a reusable plotting function. The function aggregates taxa at a selected rank, retains the most abundant groups, combines the remainder as 'Other', and creates faceted stacked-bar plots.

```{r}
# Taxonomic composition plots

# 1. Prepare object for taxonomic plots
# ps_tss should already be filtered and transformed to relative abundance
ps_tax_plot <- ps_tss

# Optional: check ranks
rank_names(ps_tax_plot)

# Make sure treatment is ordered
sample_data(ps_tax_plot)$treatment <- factor(
  sample_data(ps_tax_plot)$treatment,
  levels = c("C", "L", "H")
)

# ------------------------------------------------------------
# 2. Function for composition plots
# ------------------------------------------------------------

make_tax_composition_plot <- function(ps_obj,
                                      tax_rank = "Phylum",
                                      top_n = 10,
                                      x_var = "sample",
                                      facet_var = "treatment") {
  
  # Agglomerate taxa at selected rank
  ps_glom <- tax_glom(ps_obj, taxrank = tax_rank, NArm = FALSE)
  
  # Convert to long format
  tax_df <- psmelt(ps_glom)
  
  # Rename missing/unclassified taxa
  tax_df[[tax_rank]] <- as.character(tax_df[[tax_rank]])
  tax_df[[tax_rank]][is.na(tax_df[[tax_rank]]) | tax_df[[tax_rank]] == ""] <- "Unclassified"
  
  # Identify most abundant taxa
  top_taxa <- tax_df %>%
    group_by(.data[[tax_rank]]) %>%
    summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abundance)) %>%
    slice_head(n = top_n) %>%
    pull(.data[[tax_rank]])
  
  # Collapse low-abundance taxa into Other
  tax_df <- tax_df %>%
    mutate(
      Taxon = ifelse(.data[[tax_rank]] %in% top_taxa,
                     .data[[tax_rank]],
                     "Other"),
      Taxon = factor(Taxon)
    )
  
  # Order taxa by mean abundance, keeping Other last
  tax_order <- tax_df %>%
    group_by(Taxon) %>%
    summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abundance)) %>%
    pull(Taxon)
  
  tax_order <- c(setdiff(tax_order, "Other"), "Other")
  
  tax_df$Taxon <- factor(tax_df$Taxon, levels = rev(tax_order))
  
  # Plot
  ggplot(tax_df, aes(x = .data[[x_var]], y = Abundance, fill = Taxon)) +
    geom_bar(stat = "identity", width = 0.95) +
    facet_grid(
      cols = vars(.data[[facet_var]]),
      scales = "free_x",
      space = "free_x"
    ) +
    scale_y_continuous(
      labels = percent_format(accuracy = 1),
      expand = c(0, 0)
    ) +
    labs(
      x = "Samples",
      y = "Relative abundance",
      fill = tax_rank
    ) +
    theme_bw(base_size = 14) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title = element_text(face = "bold", colour = "black"),
      axis.text.y = element_text(colour = "black"),
      strip.text = element_text(face = "bold", colour = "black"),
      strip.background = element_rect(fill = "white", colour = "black"),
      legend.title = element_text(face = "bold", colour = "black"),
      legend.text = element_text(colour = "black"),
      legend.position = "right"
    )
}
```


### Use the reusable composition-plot function to display sample-level bacterial composition at the phylum level. The ten most abundant phyla are shown.
```{r}
# ------------------------------------------------------------
# 3.1. Phylum-level taxonomic composition
# ------------------------------------------------------------

p_phylum <- make_tax_composition_plot(
  ps_obj = ps_tax_plot,
  tax_rank = "Phylum",
  top_n = 10,
  x_var = "Sample",
  facet_var = "treatment"
)

p_phylum

ggsave(
  filename = here("figures", "taxonomic_composition_phylum.pdf"),
  plot = p_phylum,
  width = 10,
  height = 6
)

ggsave(
  filename = here("figures", "taxonomic_composition_phylum.jpg"),
  plot = p_phylum,
  width = 10,
  height = 6,
  dpi = 600
)
```

### Use the reusable composition-plot function to display sample-level bacterial composition at the phylum level. 

```{r}

# Mean taxonomic composition per treatment (PHYLUM)


# Prepare phylum-level data

ps_phylum <- tax_glom(ps_tss, taxrank = "Phylum", NArm = FALSE)

phylum_df <- psmelt(ps_phylum) %>%
  mutate(
    Phylum = as.character(Phylum),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", Phylum),
    Phylum = gsub("^p__", "", Phylum),
    
    Phylum = case_when(
      Phylum %in% c("Firmicutes_A", "Firmicutes_B", "Firmicutes_C") ~ "Firmicutes",
      Phylum %in% c("Bacteroidetes") ~ "Bacteroidota",
      TRUE ~ Phylum
    ),
    
    treatment = factor(
      treatment,
      levels = c("C", "L", "H"),
      labels = c("Control", "Low", "High")
    )
  )

# Identify dominant phyla

top_phyla <- phylum_df %>%
  group_by(Phylum) %>%
  summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_abundance)) %>%
  slice_head(n = 6) %>%
  pull(Phylum)

# Collapse rare phyla into Other at sample level

phylum_sample_summary <- phylum_df %>%
  mutate(
    Phylum_plot = ifelse(Phylum %in% top_phyla, Phylum, "Other")
  ) %>%
  group_by(treatment, Sample, Phylum_plot) %>%
  summarise(
    Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Check that each sample sums to 1
phylum_sample_summary %>%
  group_by(treatment, Sample) %>%
  summarise(total = sum(Abundance), .groups = "drop") %>%
  summarise(
    min_total = min(total),
    max_total = max(total)
  )

# Calculate mean relative abundance per treatment

phylum_summary <- phylum_sample_summary %>%
  group_by(treatment, Phylum_plot) %>%
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Check treatment bars sum to 1
phylum_summary %>%
  group_by(treatment) %>%
  summarise(total = sum(mean_abundance), .groups = "drop")

# Order phyla

phyla_order <- phylum_summary %>%
  group_by(Phylum_plot) %>%
  summarise(global_mean = mean(mean_abundance), .groups = "drop") %>%
  arrange(desc(global_mean)) %>%
  pull(Phylum_plot)

phyla_order <- c(setdiff(phyla_order, "Other"), "Other")

phylum_summary <- phylum_summary %>%
  mutate(
    Phylum_plot = factor(Phylum_plot, levels = phyla_order)
  )

# Colours

# Fixed phylum palette avoiding treatment colours
phylum_cols <- c(
  "Proteobacteria"   = "#4C78A8",  # blue
  "Firmicutes"       = "#F58518",  # orange
  "Bacteroidota"     = "#B279A2",  # purple
  "Actinobacteriota" = "#72B7B2",  # teal
  "Cyanobacteria"    = "#A65628",  # brown
  "Asgardarchaeota"  = "#E6A0C4",  # pink
  "Other"            = "#BDBDBD"   # grey
)

# Keep only colours for phyla present in the plot
phylum_cols <- phylum_cols[levels(phylum_summary$Phylum_plot)]

# Safety check: stop if any phylum is missing a colour
if (any(is.na(phylum_cols))) {
  missing_cols <- levels(phylum_summary$Phylum_plot)[is.na(phylum_cols)]
  stop(
    "Missing colours for: ",
    paste(missing_cols, collapse = ", ")
  )
}

# Plot

p_phylum_mean <- ggplot(
  phylum_summary,
  aes(x = treatment, y = mean_abundance, fill = Phylum_plot)
) +
  geom_bar(
    stat = "identity",
    position = "stack",
    width = 0.7,
    colour = "white",
    linewidth = 0.2
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.03))
  ) +
  scale_fill_manual(values = phylum_cols) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = "Treatment",
    y = "Mean relative abundance",
    fill = "Phylum"
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold", colour = "black"),
    axis.text = element_text(colour = "black"),
    legend.title = element_text(face = "bold", colour = "black"),
    legend.text = element_text(colour = "black", size = 10),
    legend.key.size = unit(0.6, "cm"),
    legend.position = "right"
  )

p_phylum_mean

# Save figure

ggsave(
  filename = here("figures", "phylum_mean_composition_final.pdf"),
  plot = p_phylum_mean,
  width = 6,
  height = 5
)

ggsave(
  filename = here("figures", "phylum_mean_composition_final.jpg"),
  plot = p_phylum_mean,
  width = 6,
  height = 5,
  dpi = 600
)


```
### Summarise genus-level relative abundance by treatment using the same general workflow as the phylum analysis. 

```{r}
# Genus-level taxonomic composition

# Prepare genus-level data

ps_genus <- tax_glom(ps_tss, taxrank = "Genus", NArm = FALSE)

genus_df <- psmelt(ps_genus) %>%
  mutate(
    Genus = as.character(Genus),
    Genus = ifelse(is.na(Genus) | Genus == "", "Unclassified", Genus),
    Genus = gsub("^g__", "", Genus),
    
    treatment = factor(
      treatment,
      levels = c("C", "L", "H"),
      labels = c("Control", "Low", "High")
    )
  )

# Identify dominant genera

top_genera <- genus_df %>%
  group_by(Genus) %>%
  summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_abundance)) %>%
  slice_head(n = 8) %>%
  pull(Genus)

# Collapse rare genera into Other at sample level

genus_sample_summary <- genus_df %>%
  mutate(
    Genus_plot = ifelse(Genus %in% top_genera, Genus, "Other")
  ) %>%
  group_by(treatment, Sample, Genus_plot) %>%
  summarise(
    Abundance = sum(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Check that each sample sums to 1
genus_sample_summary %>%
  group_by(treatment, Sample) %>%
  summarise(total = sum(Abundance), .groups = "drop") %>%
  summarise(
    min_total = min(total),
    max_total = max(total)
  )

# Calculate mean relative abundance per treatment

genus_summary <- genus_sample_summary %>%
  group_by(treatment, Genus_plot) %>%
  summarise(
    mean_abundance = mean(Abundance, na.rm = TRUE),
    .groups = "drop"
  )

# Check treatment bars sum to 1
genus_summary %>%
  group_by(treatment) %>%
  summarise(total = sum(mean_abundance), .groups = "drop")

# Order genera by global abundance, with Other last

genus_order <- genus_summary %>%
  group_by(Genus_plot) %>%
  summarise(global_mean = mean(mean_abundance), .groups = "drop") %>%
  arrange(desc(global_mean)) %>%
  pull(Genus_plot)

genus_order <- c(setdiff(genus_order, "Other"), "Other")

genus_summary <- genus_summary %>%
  mutate(
    Genus_plot = factor(Genus_plot, levels = genus_order)
  )

# Manual colour palette

n_genera <- length(levels(genus_summary$Genus_plot))

genus_cols <- scales::hue_pal()(n_genera)
names(genus_cols) <- levels(genus_summary$Genus_plot)

if ("Pseudomonas" %in% names(genus_cols)) {
  genus_cols["Pseudomonas"] <- "#A6CEE3"
}

if ("Other" %in% names(genus_cols)) {
  genus_cols["Other"] <- "grey85"
}

# Plot

p_genus_mean <- ggplot(
  genus_summary,
  aes(x = treatment, y = mean_abundance, fill = Genus_plot)
) +
  geom_bar(
    stat = "identity",
    position = "stack",
    width = 0.7,
    colour = "white",
    linewidth = 0.2
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.03))
  ) +
  scale_fill_manual(values = genus_cols) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = "Treatment",
    y = "Mean relative abundance",
    fill = "Genus"
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold", colour = "black"),
    axis.text = element_text(colour = "black"),
    legend.title = element_text(face = "bold", colour = "black"),
    legend.text = element_text(colour = "black", size = 10),
    legend.key.size = unit(0.6, "cm"),
    legend.position = "right"
  )

p_genus_mean

# Save figure

ggsave(
  filename = here("figures", "genus_mean_composition_final.pdf"),
  plot = p_genus_mean,
  width = 7,
  height = 5
)

ggsave(
  filename = here("figures", "genus_mean_composition_final.jpg"),
  plot = p_genus_mean,
  width = 7,
  height = 5,
  dpi = 600
)

```

### Construct the final beta-diversity ordination using principal coordinates analysis of Aitchison distances.

```{r}
# Beta diversity: PCoA (CLR-transformed data)

# CLR transformation

ps_clr <- microbiome::transform(ps_NB, "clr")

# Distance matrix (Aitchison = Euclidean on CLR)

clr_mat <- as(otu_table(ps_clr), "matrix")

# Ensure taxa are columns
if (taxa_are_rows(ps_clr)) {
  clr_mat <- t(clr_mat)
}

clr_dist <- dist(clr_mat, method = "euclidean")

# PCoA

pcoa_res <- cmdscale(clr_dist, eig = TRUE, k = 2)

# Extract coordinates
pcoa_df <- as.data.frame(pcoa_res$points)
colnames(pcoa_df) <- c("PCoA1", "PCoA2")

# Add metadata

meta_df <- data.frame(sample_data(ps_clr))

pcoa_df <- pcoa_df %>%
  mutate(
    Sample = rownames(pcoa_df)
  ) %>%
  left_join(
    meta_df %>% rownames_to_column("Sample"),
    by = "Sample"
  )

# Format treatment
pcoa_df$treatment <- factor(
  pcoa_df$treatment,
  levels = c("C", "L", "H"),
  labels = c("Control", "Low", "High")
)

# Variance explained

eig_vals <- pcoa_res$eig

var_explained <- eig_vals / sum(eig_vals)

pc1_var <- round(var_explained[1] * 100, 1)
pc2_var <- round(var_explained[2] * 100, 1)

# Plot

p_pcoa <- ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, colour = treatment)) +
  geom_point(size = 3, alpha = 0.8) +
  
  # Ellipses (visual grouping only)
  stat_ellipse(
    aes(group = treatment),
    type = "t",
    linetype = 2,
    linewidth = 0.8
  ) +
  
  scale_colour_manual(values = c(
    "Control" = "#2E8B57",
    "Low" = "#C9A227",
    "High" = "#B22222"
  )) +
  
  labs(
    x = paste0("PCoA1 (", pc1_var, "%)"),
    y = paste0("PCoA2 (", pc2_var, "%)"),
    colour = "Treatment"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(face = "bold", colour = "black"),
    axis.text = element_text(colour = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

p_pcoa

# Save figure

ggsave(
  filename = here("figures", "pcoa_clr.pdf"),
  plot = p_pcoa,
  width = 6,
  height = 5
)

ggsave(
  filename = here("figures", "pcoa_clr.jpg"),
  plot = p_pcoa,
  width = 6,
  height = 5,
  dpi = 600
)
```

```{r}
# Merge figures
figure_microbiome <- (
  p_pcoa +
  p_phylum_mean +
  plot_layout(widths = c(1.4, 1.6))
) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(
      size = 18,
      face = "bold"
    )
  )

figure_microbiome

ggsave(
  filename = here("figures", "beta_phylum.pdf"),
  plot = figure_microbiome,
  width = 12,
  height = 5
)

ggsave(
  filename = here("figures", "beta_phylum.jpg"),
  plot = figure_microbiome,
  width = 12,
  height = 5,
  dpi = 600
)
```

###########################################################
###########################################################

## Sequencing depth quality control

### Create a helper function for extracting sample sequencing depths from phyloseq objects, then compare read counts before and after filtering. The resulting table quantifies reads removed and the proportion retained for every sample.

```{r}
# Helper function: extract sample read depths

get_sample_depths <- function(ps_object, depth_name) {

  df <- data.frame(
    sample = sample_names(ps_object),
    reads = sample_sums(ps_object)
  )

  colnames(df)[2] <- depth_name

  df
}

# Read depth before taxonomic/prevalence filtering
depth_raw <- get_sample_depths(
  ps_object = ps_NB_raw,
  depth_name = "reads_before_filtering"
)

depth_filtered <- get_sample_depths(
  ps_object = ps_NB,
  depth_name = "reads_after_filtering"
)

# Combine before and after read depths
depth_comparison <- full_join(
  depth_raw,
  depth_filtered,
  by = "sample"
) %>%
  mutate(
    reads_removed = reads_before_filtering - reads_after_filtering,
    proportion_retained = reads_after_filtering / reads_before_filtering,
    percent_retained = 100 * proportion_retained,
    percent_removed = 100 - percent_retained
  ) %>%
  arrange(reads_after_filtering)

depth_comparison
```

### Summarise the dimensions and sequencing depth of the microbiome dataset before and after filtering. 

```{r}
# Number of samples before and after filtering
sample_summary <- tibble(
  dataset = c("Before filtering", "After filtering"),
  n_samples = c(nsamples(ps_NB_raw), nsamples(ps_NB)),
  n_ASVs = c(ntaxa(ps_NB_raw), ntaxa(ps_NB)),
  total_reads = c(
    sum(sample_sums(ps_NB_raw)),
    sum(sample_sums(ps_NB))
  ),
  mean_reads = c(
    mean(sample_sums(ps_NB_raw)),
    mean(sample_sums(ps_NB))
  ),
  median_reads = c(
    median(sample_sums(ps_NB_raw)),
    median(sample_sums(ps_NB))
  ),
  minimum_reads = c(
    min(sample_sums(ps_NB_raw)),
    min(sample_sums(ps_NB))
  ),
  maximum_reads = c(
    max(sample_sums(ps_NB_raw)),
    max(sample_sums(ps_NB))
  )
)

sample_summary
```

### Count how many filtered samples fall below several candidate sequencing-depth thresholds. 
```{r}
# Thresholds chosen to show the lower tail of the distribution
depth_thresholds <- c(1000, 2000, 3000, 5000, 10000)

low_depth_counts <- tibble(
  threshold = depth_thresholds,
  n_samples_below_threshold = sapply(
    depth_thresholds,
    function(x) sum(sample_sums(ps_NB) < x)
  ),
  percent_samples_below_threshold = sapply(
    depth_thresholds,
    function(x) {
      100 * mean(sample_sums(ps_NB) < x)
    }
  )
)

low_depth_counts
```

### Apply a 5,000-read diagnostic threshold and list all samples below it. This step is exploratory and identifies samples that may warrant closer inspection rather than automatically removing them.

```{r}
sum(sample_sums(ps_NB) < 5000)

low_depth_cutoff <- 5000

low_depth_samples <- depth_comparison %>%
  filter(reads_after_filtering < low_depth_cutoff) %>%
  arrange(reads_after_filtering)

low_depth_samples
```

### Join the low-depth sample list to the experimental metadata. 

```{r}
metadata_df <- data.frame(sample_data(ps_NB)) %>%
  rownames_to_column("sample")

low_depth_samples_metadata <- low_depth_samples %>%
  left_join(metadata_df, by = "sample")

low_depth_samples_metadata
```

### Plot the distribution of post-filtering read depths and mark the 5,000-read diagnostic cutoff. 

```{r}
ggplot(depth_comparison, aes(x = reads_after_filtering)) +
  geom_histogram(bins = 30, colour = "black", fill = "grey80") +
  geom_vline(
    xintercept = low_depth_cutoff,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  labs(
    x = "Reads per sample after filtering",
    y = "Number of samples"
  ) +
  theme_bw()
```

### Compare each sample's read depth before and after filtering.

```{r}
ggplot(
  depth_comparison,
  aes(
    x = reads_before_filtering,
    y = reads_after_filtering
  )
) +
  geom_point() +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed"
  ) +
  labs(
    x = "Reads before filtering",
    y = "Reads after filtering"
  ) +
  theme_bw()
```

```{r}
low_depth_loss_summary <- low_depth_samples_metadata %>%
  select(
    sample,
    reads_before_filtering,
    reads_after_filtering,
    reads_removed,
    percent_retained,
    percent_removed,
    everything()
  )

low_depth_loss_summary
```



```{r}
# DETAILED QC REPORT FOR EXTREME READ-LOSS SAMPLES
# Samples identified as having unusually high read loss
samples_to_check <- c(
  "60HT17.fastq",
  "69HT8.fastq",
  "81CT9.fastq"
)

# ------------------------------------------------------------
# 1. Helper: convert phyloseq OTU table to samples x ASVs
# ------------------------------------------------------------

otu_samples_by_taxa <- function(ps_object) {

  otu <- as(otu_table(ps_object), "matrix")

  if (taxa_are_rows(ps_object)) {
    otu <- t(otu)
  }

  return(otu)
}

otu_raw <- otu_samples_by_taxa(ps_NB_raw)
otu_filtered <- otu_samples_by_taxa(ps_NB)

# Keep only samples present in both objects
samples_to_check <- intersect(
  samples_to_check,
  intersect(rownames(otu_raw), rownames(otu_filtered))
)

samples_to_check
```

```{r}
read_loss_report <- tibble(
  sample = samples_to_check,
  reads_before = rowSums(otu_raw[samples_to_check, , drop = FALSE]),
  reads_after = rowSums(otu_filtered[samples_to_check, , drop = FALSE])
) %>%
  mutate(
    reads_removed = reads_before - reads_after,
    percent_retained = 100 * reads_after / reads_before,
    percent_removed = 100 - percent_retained
  ) %>%
  arrange(percent_retained)

read_loss_report
```
### Quantify total reads before and after filtering for each flagged sample. 

```{r}
taxonomy_raw <- as(tax_table(ps_NB_raw), "matrix") %>%
  as.data.frame() %>%
  rownames_to_column("ASV")

taxonomy_filtered <- as(tax_table(ps_NB), "matrix") %>%
  as.data.frame() %>%
  rownames_to_column("ASV")

get_top_ASVs_before <- function(sample_name, n_top = 20) {

  counts <- otu_raw[sample_name, ]

  tibble(
    sample = sample_name,
    ASV = names(counts),
    reads_before = as.numeric(counts)
  ) %>%
    filter(reads_before > 0) %>%
    mutate(
      relative_abundance_before =
        reads_before / sum(reads_before)
    ) %>%
    arrange(desc(reads_before)) %>%
    slice_head(n = n_top) %>%
    left_join(taxonomy_raw, by = "ASV") %>%
    mutate(
      retained_after_filtering =
        ASV %in% colnames(otu_filtered)
    )
}
top_ASVs_before <- bind_rows(
  lapply(
    samples_to_check,
    get_top_ASVs_before,
    n_top = 20
  )
)

top_ASVs_before

top_ASVs_before %>%
  select(
    sample,
    ASV,
    reads_before,
    relative_abundance_before,
    retained_after_filtering,
    Domain,
    Phylum,
    Class,
    Order,
    Family,
    Genus,
    Species
  )
```

### Retrieve the post-filtering count for each dominant ASV identified in the flagged samples. 

```{r}
get_after_count <- function(sample_name, ASV_name) {

  if (
    sample_name %in% rownames(otu_filtered) &&
    ASV_name %in% colnames(otu_filtered)
  ) {
    return(otu_filtered[sample_name, ASV_name])
  }

  return(0)
}

top_ASVs_comparison <- top_ASVs_before %>%
  rowwise() %>%
  mutate(
    reads_after = get_after_count(sample, ASV),
    reads_removed = reads_before - reads_after
  ) %>%
  ungroup() %>%
  mutate(
    percent_of_ASV_reads_retained =
      100 * reads_after / reads_before
  )

top_ASVs_comparison %>%
  select(
    sample,
    ASV,
    reads_before,
    reads_after,
    reads_removed,
    relative_abundance_before,
    retained_after_filtering,
    percent_of_ASV_reads_retained,
    Phylum,
    Class,
    Order,
    Family,
    Genus
  )
```




```{r}
ASV_dataset_context <- tibble(
  ASV = colnames(otu_raw),
  total_reads_all_samples = colSums(otu_raw),
  prevalence_n_samples = colSums(otu_raw > 0),
  prevalence_percent =
    100 * colSums(otu_raw > 0) / nrow(otu_raw),
  mean_reads_when_present = apply(
    otu_raw,
    2,
    function(x) {
      if (sum(x > 0) == 0) return(0)
      mean(x[x > 0])
    }
  ),
  median_reads_when_present = apply(
    otu_raw,
    2,
    function(x) {
      if (sum(x > 0) == 0) return(0)
      median(x[x > 0])
    }
  )
)

top_ASVs_context <- top_ASVs_comparison %>%
  left_join(
    ASV_dataset_context,
    by = "ASV"
  ) %>%
  arrange(sample, desc(reads_before))

top_ASVs_context %>%
  select(
    sample,
    ASV,
    reads_before,
    reads_after,
    relative_abundance_before,
    retained_after_filtering,
    prevalence_n_samples,
    prevalence_percent,
    total_reads_all_samples,
    Phylum,
    Class,
    Order,
    Family,
    Genus
  )
```

### Run basic integrity checks on the final filtered phyloseq object before differential-abundance analysis. The checks confirm sample and ASV counts, sequencing-depth ranges, and the presence of the treatment and tank metadata variables required by the model.

```{r}
# BASIC CHECKS

ps_NB

nsamples(ps_NB)
ntaxa(ps_NB)

summary(sample_sums(ps_NB))
range(sample_sums(ps_NB))

# Inspect metadata column names
colnames(sample_data(ps_NB))

# Confirm treatment and tank are present
stopifnot(
  "treatment" %in% colnames(sample_data(ps_NB)),
  "tank" %in% colnames(sample_data(ps_NB))
)
```

```{r}
ps_NB <- prune_taxa(
  taxa_sums(ps_NB) > 0,
  ps_NB
)
```

### Prepare raw count data for genus-level MaAsLin2 analysis. 

```{r}
# AGGREGATE RAW COUNTS TO GENUS LEVEL

# Remove taxa without a usable genus assignment
ps_genus_maaslin <- subset_taxa(
  ps_NB,
  !is.na(Genus) &
    Genus != "" &
    Genus != "g__Unclassified" &
    Genus != "Unclassified"
)

# Agglomerate ASVs belonging to the same genus
ps_genus_maaslin <- tax_glom(
  ps_genus_maaslin,
  taxrank = "Genus",
  NArm = TRUE
)

# Remove taxa with zero counts, if any
ps_genus_maaslin <- prune_taxa(
  taxa_sums(ps_genus_maaslin) > 0,
  ps_genus_maaslin
)

ps_genus_maaslin

nsamples(ps_genus_maaslin)
ntaxa(ps_genus_maaslin)
```

```{r}
# EXTRACT GENUS COUNT TABLE

genus_counts <- as(
  otu_table(ps_genus_maaslin),
  "matrix"
)

# Ensure samples are rows and genera are columns
if (taxa_are_rows(ps_genus_maaslin)) {
  genus_counts <- t(genus_counts)
}

genus_counts <- as.data.frame(genus_counts)

dim(genus_counts)
genus_counts[1:5, 1:min(5, ncol(genus_counts))]
```

```{r}
# ASSIGN READABLE GENUS NAMES

genus_taxonomy <- as.data.frame(
  tax_table(ps_genus_maaslin)
)

genus_taxonomy$Genus <- as.character(genus_taxonomy$Genus)

# Remove GTDB-style genus prefix
genus_taxonomy$Genus <- gsub(
  "^g__",
  "",
  genus_taxonomy$Genus
)

# Ensure names are syntactically valid and unique
genus_names <- make.unique(
  make.names(genus_taxonomy$Genus)
)

# Taxonomy rows correspond to genus-count columns
colnames(genus_counts) <- genus_names

# Confirm unique names
anyDuplicated(colnames(genus_counts))
# Should return 0

head(colnames(genus_counts))
```

```{r}
# PREPARE SAMPLE METADATA

maaslin_metadata <- data.frame(
  sample_data(ps_genus_maaslin)
)

# Ensure row names are sample IDs
head(rownames(maaslin_metadata))
head(rownames(genus_counts))

# Add sequencing depth based on the final filtered ASV object
maaslin_metadata$reads <- sample_sums(ps_NB)[
  rownames(maaslin_metadata)
]

maaslin_metadata$log_reads <- log(
  maaslin_metadata$reads
)

# Treatment: set Control as reference
maaslin_metadata$treatment <- factor(
  maaslin_metadata$treatment,
  levels = c("C", "L", "H"),
  labels = c("Control", "Low", "High")
)

# Tank must be categorical
maaslin_metadata$tank <- factor(
  maaslin_metadata$tank
)

# Check treatment sample sizes
table(maaslin_metadata$treatment)

# Check tank replication
table(maaslin_metadata$tank)

# Mean number of fish per tank
mean(table(maaslin_metadata$tank))
```

```{r}
# ALIGN SAMPLE ORDER

common_samples <- intersect(
  rownames(genus_counts),
  rownames(maaslin_metadata)
)

length(common_samples)

genus_counts <- genus_counts[
  common_samples,
  ,
  drop = FALSE
]

maaslin_metadata <- maaslin_metadata[
  common_samples,
  ,
  drop = FALSE
]

# Put metadata in exactly the same order
maaslin_metadata <- maaslin_metadata[
  rownames(genus_counts),
  ,
  drop = FALSE
]

# Confirm perfect alignment
stopifnot(
  identical(
    rownames(genus_counts),
    rownames(maaslin_metadata)
  )
)

dim(genus_counts)
dim(maaslin_metadata)
```

```{r}
# CHECK GENUS PREVALENCE AND ABUNDANCE BEFORE MAASLIN

# Relative abundance only for QC summaries
genus_relative_abundance <- sweep(
  as.matrix(genus_counts),
  1,
  rowSums(genus_counts),
  "/"
)

genus_prevalence <- colMeans(
  genus_counts > 0
)

genus_mean_abundance <- colMeans(
  genus_relative_abundance,
  na.rm = TRUE
)

genus_filter_summary <- tibble(
  Genus = colnames(genus_counts),
  prevalence = genus_prevalence,
  prevalence_percent = genus_prevalence * 100,
  mean_relative_abundance = genus_mean_abundance
) %>%
  arrange(desc(mean_relative_abundance))

genus_filter_summary
```



```{r}
# OUTPUT DIRECTORY

maaslin_output <- here(
  "outputs",
  "maaslin2_genus"
)

# Remove an old output directory before rerunning
# Uncomment only when you deliberately want to overwrite it
#
# if (dir.exists(maaslin_output)) {
#   unlink(maaslin_output, recursive = TRUE)
# }

dir.create(
  maaslin_output,
  recursive = TRUE,
  showWarnings = FALSE
)
```


### Fit the genus-level MaAsLin2 model. Treatment and sequencing depth are included as fixed effects, tank is included as a random effect, Control is the reference group, and multiple testing is controlled using the Benjamini-Hochberg false-discovery-rate procedure.

```{r}
set.seed(123)

fit_maaslin2_genus <- Maaslin2::Maaslin2(
  input_data = genus_counts,
  input_metadata = maaslin_metadata,
  output = maaslin_output,

  fixed_effects = c(
    "treatment",
    "log_reads"
  ),

  random_effects = c(
    "tank"
  ),

  reference = "treatment,Control",

  normalization = "TSS",
  transform = "LOG",
  analysis_method = "LM",

  min_prevalence = 0.10,
  min_abundance = 0.0001,

  correction = "BH",
  standardize = FALSE,

  max_significance = 0.05,
  plot_heatmap = TRUE,
  plot_scatter = TRUE,

  save_models = TRUE
)
```


```{r}
maaslin_all_results <- read_tsv(
  file.path(
    maaslin_output,
    "all_results.tsv"
  ),
  show_col_types = FALSE
)

glimpse(maaslin_all_results)

maaslin_all_results %>%
  arrange(qval) %>%
  head(20)
```

```{r}
treatment_results <- maaslin_all_results %>%
  filter(metadata == "treatment") %>%
  arrange(qval, pval)

treatment_results
```


```{r}
unique(treatment_results$value)

```

### Identify treatment-associated genera that remain significant after false-discovery-rate correction at q < 0.05. 
```{r}
significant_treatment_results <- treatment_results %>%
  filter(qval < 0.05) %>%
  arrange(qval)

significant_treatment_results

nrow(significant_treatment_results)
```

### Add effect-direction and approximate fold-change interpretations to the treatment results. Positive coefficients indicate higher abundance than Control, whereas negative coefficients indicate lower abundance.

```{r}
treatment_results_interpreted <- treatment_results %>%
  mutate(
    approximate_fold_change = exp(coef),
    direction = case_when(
      coef > 0 ~ "Higher than Control",
      coef < 0 ~ "Lower than Control",
      TRUE ~ "No change"
    )
  ) %>%
  arrange(qval)

treatment_results_interpreted %>%
  select(
    feature,
    value,
    coef,
    approximate_fold_change,
    direction,
    pval,
    qval,
    N,
    N.not.0
  )
```

```{r}
genus_long <- genus_counts %>%
  rownames_to_column("sample") %>%
  pivot_longer(
    cols = -sample,
    names_to = "feature",
    values_to = "count"
  ) %>%
  left_join(
    maaslin_metadata %>%
      rownames_to_column("sample") %>%
      select(
        sample,
        treatment,
        tank,
        reads
      ),
    by = "sample"
  ) %>%
  mutate(
    detected = count > 0
  )

genus_support <- genus_long %>%
  group_by(
    feature,
    treatment
  ) %>%
  summarise(
    n_samples = n(),
    n_detected = sum(detected),
    prevalence_percent = 100 * mean(detected),
    median_count_when_present = ifelse(
      any(detected),
      median(count[detected]),
      NA_real_
    ),
    .groups = "drop"
  )

genus_support
```

```{r}
significant_genus_support <- genus_support %>%
  filter(
    feature %in%
      significant_treatment_results$feature
  )

significant_genus_support
```

```{r}
if (nrow(significant_treatment_results) > 0) {

  significant_genera <- unique(
    significant_treatment_results$feature
  )

  plot_data_significant <- genus_long %>%
    filter(feature %in% significant_genera) %>%
    group_by(sample) %>%
    mutate(
      relative_abundance =
        count / sum(count)
    ) %>%
    ungroup()

  p_maaslin_significant <- ggplot(
    plot_data_significant,
    aes(
      x = treatment,
      y = relative_abundance
    )
  ) +
    geom_violin(
      trim = FALSE,
      alpha = 0.3
    ) +
    geom_jitter(
      width = 0.15,
      alpha = 0.6
    ) +
    facet_wrap(
      ~ feature,
      scales = "free_y"
    ) +
    scale_y_continuous(
      labels = scales::percent_format()
    ) +
    labs(
      x = "Fluoxetine treatment",
      y = "Relative abundance"
    ) +
    theme_bw()

  p_maaslin_significant

  ggsave(
    filename = here(
      "figures",
      "maaslin2_significant_genera.pdf"
    ),
    plot = p_maaslin_significant,
    width = 9,
    height = 6
  )
}
```

```{r}
genus_prevalence_summary <- tibble(
  Genus = colnames(genus_counts),
  n_samples_present = colSums(genus_counts > 0),
  prevalence = colMeans(genus_counts > 0),
  prevalence_percent = 100 * colMeans(genus_counts > 0),
  total_reads = colSums(genus_counts)
) %>%
  arrange(desc(prevalence))
```

### Export the principal MaAsLin2 result and quality-control tables. Separate files are created for all treatment results, FDR-significant results, genus prevalence, and treatment-specific support summaries.

```{r}
write_csv(
  treatment_results_interpreted,
  file.path(
    maaslin_output,
    "treatment_results_all.csv"
  )
)

write_csv(
  significant_treatment_results,
  file.path(
    maaslin_output,
    "treatment_results_significant_q005.csv"
  )
)

write_csv(
  genus_prevalence_summary,
  file.path(
    maaslin_output,
    "genus_prevalence_summary.csv"
  )
)

write_csv(
  genus_support,
  file.path(
    maaslin_output,
    "genus_support_by_treatment.csv"
  )
)
```

### Summarise the differential-abundance analysis by reporting the number of genera tested and the number of treatment associations significant after FDR correction. 

```{r}
n_genera_tested <- treatment_results %>%
  distinct(feature) %>%
  nrow()

n_significant <- significant_treatment_results %>%
  distinct(feature, value) %>%
  nrow()

cat(
  "Genera tested:", n_genera_tested, "\n",
  "Significant treatment associations after FDR correction:",
  n_significant, "\n"
)
```
```{r}
table(maaslin_metadata$treatment)

nrow(treatment_results)

significant_treatment_results

treatment_results_interpreted %>%
  select(
    feature,
    value,
    coef,
    pval,
    qval,
    N,
    N.not.0
  ) %>%
  head(30)
```


